---
title: "<b>Apply Family Notes</b>"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r setup, eval = TRUE, include=FALSE}

# Setting global options for the document
knitr::opts_chunk$set( )
```

```{r packages, include = FALSE, echo=FALSE}
#Installing/loading required packages
`Required Packages` <- c("tidyverse","knitr","kableExtra") 
`New Packages` <- `Required Packages`[!(`Required Packages` %in% installed.packages()[,"Package"])]
if(length(`New Packages`)) install.packages(`New Packages`, repos = "http://cran.us.r-project.org")
invisible(suppressPackageStartupMessages(suppressWarnings(lapply(`Required Packages`, require, character.only=T))))
```

#   **Purpose**

In this notebook session, I'll be going over the "Apply" Family in base R. The "Apply Functions" refer to a group of functions that come with base R that allow you to do repetitive actions within different objects(i.e. data frames, lists, etc.)

The functions I'll go over will be:

-   apply()
-   lapply()
-   mapply()
-   rapply()
-   sapply()
-   tapply()
-   vapply()

# **Apply Functions, Inputs, & Outputs**

A quick overview:

```{r applychartwork, echo=F}

Name = c("apply()","lapply()","mapply()","rapply()","sapply()","tapply()","vapply()")
`What it does` = c("Applies a function to the rows or columns of the object", "Applies a function to all elements within the input","Applies a function to multiple lists or vectors - can be considered a multivariate version of 'sapply'", "Applies a function recursively through a list - nested lists", "A simpler version of 'lapply' that works on list, data frames, and vectors","Applies a function over a ragged/jagged array (an array that has more than one dimension with varying lengths)","Similar to 'sapply', but you can pre-specify the type of value that is output, making it a bit faster")
Input = c("Data Frame or Matrix","Data Frame, List, Vector","Multiple Lists or Vectors (i.e. a Data Frame)","Nested Lists","Data Frame, List, Vector","Data Frame or Vector that can be split (divided into groups/factors)","Data Frame, List, Vector")
Output = c("Matrix or Array","List","List or Vector","Nested List or Vector depending on arguments passed","Matrix or Vector", "Array","Data Frame, List, Vector")
applychart <- tibble(Name,`What it does`,Input,Output)

kable(applychart) %>%
   kable_minimal()

```

# **Apply Function: apply()**

The **apply()** function is used to apply a function to all rows or columns of an object. Consequently, only objects with more than one dimension can be used with apply, so a data frame or matrix.


`apply(X, MARGIN, FUN)`


Where:

```{r, echo =F}
Argument <- c("X","MARGIN","FUN")
Description <- c("Data Frame or Matrix","'1' or '2' or 'c(1,2)' where 1 = Rows and 2 = Columns","The function you want to be applied to the data frame or matrix in question")
applyfunctionchart <- tibble(Argument,Description)
kable(applyfunctionchart) %>%
  kable_minimal()
```


## **Data Frame Example**

Creating a mock data frame:

```{r applydfex, include = T}

City <- c("Buffalo","NYC","Seattle","Austin","Orlando","Minneapolis")
Cases <- c(500,2012,1876,635,4512,823)
Controls <- c(3426,5210,6753,5633,2013,1890)

records <- data.frame(City,Cases,Controls, row.names = NULL)

records

```


We can use the apply function to calculate column sums...

```{r, include = T}

apply(records[,2:3], 2,sum)
```


...Or row sums. (Note that these both produce vectors)

```{r, include = T}

apply(records[,2:3], 1,sum)
```


We can name the vectors in one line of code with the `names<-` function 
```{r, include = T}

`City Totals` <-  `names<-`(apply(records[,2:3], 1,sum), records$City)

`City Totals`
```

## **Applying Statistic (More Complex) Functions**

We can also do different statistical procedures based on each test's requirement. Let's do a T-test:

```{r, include = T}

#Making a mock data set for T-test
Exam_one_grades <- c(67,53)
Exam_two_grades <- c(90,89)
Exam_three_grades <- c(89,95)
Exam_four_grades <- c(95,87)
Exam_five_grades <- c(100,99)
Student <- c("Student1","Student2")

Grades <- tibble(Student,Exam_one_grades,Exam_two_grades,Exam_three_grades,Exam_four_grades,Exam_five_grades)

Grades

```

Let's just say we want the P.value of one sample T-tests for each student and we want to place it in this dataset as a new column. What's important to note is that arguments that would normally be passed through to your functions, go as seperate arguments at the end of the apply function, after you declare which function you want used.

```{r, include = T}

#Turning off scientific notation formatting
options(scipen = 999)

#Getting index of columns that end with the word "grades"
Gradeindexes <- grep(("grades$"),names(Grades))

#Using apply to apply the t.test.
testresults <- apply(Grades[,Gradeindexes],1,t.test, alternative = "two.sided", conf.level = 0.95)

#Using do.call to bind p values to the data set. because results are in a list, need to use lapply.

Grades$`P Values` <- do.call(rbind, lapply(testresults, function(x){x$p.value}))

Grades
```

# **Lapply Function: lapply()**
TBC

# **Mapply Function: mapply()**
TBC

# **Rapply Function: rapply()**
TBC

# **Sapply Function: sapply()**
TBC

# **Tapply Function: tapply()**
TBC

# **Vapply Function: vapply**
TBC
